A cautionary note on the use of Ornstein Uhlenbeck and other models in macroevolutionary studies

### 2015-04-21 Manuscript/ohyou_suppmat.tex

\section{Supplementary Material}

@@

    For each of these papers we recorded the number of species in the analysis, the study group (amphibians, birds, fish, mammals, reptiles, invertebrates or plants), the statistical package or specific R package used to fit the models, and the reason the authors state for using an OU model (ancestral state reconstructions, detecting convergent evolution, controlling for phylogeny, selecting a model of trait evolution, or other). 
    Where papers included multiple analyses using different numbers of species we used the median number of taxa. 
    Where papers had multiple study groups, statistics/R packages or reasons for fitting OU models we counted them in each relevant category. 
    We summarise these results in Figure 1 and Table 1, and the full dataset is available below along with the full list of references.

    In total, our literature search found 3720 papers published between 2005 and 2014, and the number has increased substantially since 2005 (Fig. 1). 
    Most papers fit OU models to phylogenies with fewer than 100 taxa (mean = 166.97 $\pm$ 43.86, median = 58, Supplemental Fig. S1 and Table S1). 
    The majority of papers fit OU models using R packages, particularly GEIGER and, although other uses are becoming more common, most papers use OU models in an effort to discern the ``best'' model of trait evolution or to control for phylogenetic non-independence (Supplemental Table S1 and Fig. S1).

\begin{landscape}
\begin{center}
\begin{longtable}{p{6cm}llllll}
\caption[Literature review]
        {Details of the papers used in our literature review. For each paper we recorded the study taxon, the number of tips in the phylogeny used to fit the Ornstein Uhlenbeck (OU) model, how the authors used the OU model in the paper, and the statistical package (usually R) used to carry out the analyses. For a full reference list see below. Journal abbreviations: Am Nat = The American Naturalist, ELE = Ecology Letters, JEB = Journal of Evolutionary Biology, PRSB = Proceedings of the Royal Society B: Biological Sciences, PNAS = Proceedings of the National Academy of the USA. †Number of tips in the phylogeny the OU model was fitted to. Where there were multiple analyses in a paper we use the median number of tips.  All packages mentioned are R packages apart from BayesTraits, COMPARE, MATLAB, Mesquite, PAM, PDAP and PDTREE.}\\
\hline

\textbf{Paper}  &   Year    &   Journal &   Taxon   &   Ntips   &   Use in paper    &   Stats/R package \\

@@

Gvoždík and Damme 2006    &   2006    &   Evolution   &   amphibians  &   10  &   phylogenetic correction &   COMPARE \\

@@

Gomez and Théry 2007  &   2007    &   Am Nat  &   birds   &   40  &   model of evolution  &   OUCH    \\

@@

Arbour and López-Fernández 2013   &   2013    &   PRSB    &   fish    &   27  &   model of evolution  &   OUCH    \\

@@

Frédérich et al. 2013 &   2013    &   Am Nat  &   fish    &   208 &   model of evolution  &   OUwie   \\

@@

López-Fernández et al. 2013   &   2013    &   Evolution   &   fish    &   135 &   model of evolution  &   GEIGER  \\

@@

Pérez i de Lanuza et al. 2013 &   2013    &   JEB &   reptiles    &   42  &   ancestral state reconstruction  &   GEIGER  \\

@@

Arbour, J. H., and H. López-Fernández. 2013. Ecological variation in South American geophagine cichlids arose during an early burst of adaptive morphological and functional evolution. Proceedings of the Royal Society B: Biological Sciences 280.\\

@@

Gomez, D., and M. Théry. 2007. Simultaneous crypsis and conspicuousness in color patterns: comparative analysis of a neotropical rainforest bird community. The American Naturalist 169:S42-S61.\\

@@

Gvoždík, L., and R. V. Damme. 2006. Triturus newts defy the running-swimming dilemma. Evolution 60:2110-2121.\\

@@

Huey, R. B., C. A. Deutsch, J. J. Tewksbury, L. J. Vitt, P. E. Hertz, H. J. Álvarez Pérez, and T. Garland. 2009. Why tropical forest lizards are vulnerable to climate warming. Proceedings of the Royal Society of London Series B-Biological Sciences 276:1939-1948.\\

@@

López-Fernández, H., J. H. Arbour, K. O. Winemiller, and R. L. Honeycutt. 2013. Testing for ancient adaptive radiations in Neotropical cichlid fishes. Evolution 67:1321-1337.\\

@@

Smith, S. D., C. Ané, and D. A. Baum. 2008. The role of pollinator shifts in the floral diversification of Iochroma (Solanaceae). Evolution 62:793-806.\\

@@

Voje, K. L., A. B. Mazzarella, T. F. Hansen, K. Østbye, T. Klepaker, A. Bass, A. Herland, K. M. Bærum, F. Gregersen, and L. A. Vøllestad. 2013. Adaptation and constraint in a stickleback radiation. Journal of evolutionary biology 26:2396-2414.\\

@@

  \subsection{Bayesian simulations} % keep? or move it all to supp mat?
    For completeness, we repeated the Maximum Likelihood simulations in a Bayesian framework implemented in BayesTraits. 
    The choice of priors for OU models has not been fully explored. 
    Therefore, we used three alternative sets of priors on $\alpha$: (i) an exponential distribution with mean = 1; (ii), an exponential distribution with mean = 10, and (iii) a uniform distribution bounded at 0 and 20. 
    For all analyses we used a uniform -100 to 100 prior for $\mu$ and uniform 0 to 100 prior for $\sigma^2$. 
    We present the results from the exponential prior with mean = 10 in the main text since this is a broad, liberal prior, but we provide results derived from the other priors in the Supplementary Material (Figs. S2-S5). % add links to this
    We ran the MCMC chains for $1x10^6$ iterations, disregarding the first $1x10^4$ as burn-in. 
    Following burn-in the chains were sampled every 1000 to ensure independence of each consecutive sample. 
    Multiple independent chains were run for each analysis to ensure convergence was reached. 
    We determined fit to an OU model using Bayes factors estimated from a stepping stone sampling procedure \citep{xie2010improving}implemented in BayesTraits. 
    The marginal likelihoods of the models were calculated using a stepping stone sampler in which fifty stones were drawn from a beta distribution with (alpha = 0.4 and beta = 1). 
    Each stone was sampled for 20000 iterations (with the first 5000 iterations discarded) 
    We treated Bayes factors $>$ 2 as evidence favouring the OU model.

>>>>

\section{Supplementary Material}
\setcounter{table}{0} \renewcommand{\thetable}{S\arabic{table}}

@@

    For each of these papers we recorded the number of species in the analysis, the study group (amphibians, birds, fish, mammals, reptiles, invertebrates or plants), the statistical package or specific R package used to fit the models, and the reason the authors state for using an OU model (ancestral state reconstructions, detecting convergent evolution, controlling for phylogeny, selecting a model of trait evolution, or other). 
    Where papers included multiple analyses using different numbers of species we used the median number of taxa. 
    Where papers had multiple study groups, statistics/R packages or reasons for fitting OU models we counted them in each relevant category. 
    We summarise these results in Figures 1 and 2 and Table 1, and the full dataset is available below along with the full list of references.

    In total, our literature search found 3720 papers published between 2005 and 2014, and the number has increased substantially since 2005 (Fig. 1). 
    Most papers fit OU models to phylogenies with fewer than 100 taxa (mean = 166.97 $\pm$ 43.86, median = 58, Figure 2 and Table S1). 
    The majority of papers fit OU models using R packages, particularly GEIGER and, although other uses are becoming more common, most papers use OU models in an effort to discern the ``best'' model of trait evolution or to control for phylogenetic non-independence (Table S1).

\begin{landscape}
\begin{center}
\begin{longtable}{p{6cm}llllll}
\caption[Literature review]
        {Details of the papers used in our literature review. For each paper we recorded the study taxon, the number of tips in the phylogeny used to fit the Ornstein Uhlenbeck (OU) model, how the authors used the OU model in the paper, and the statistical package (usually R) used to carry out the analyses. For a full reference list see below. Journal abbreviations: Am Nat = The American Naturalist, ELE = Ecology Letters, JEB = Journal of Evolutionary Biology, PRSB = Proceedings of the Royal Society B: Biological Sciences, PNAS = Proceedings of the National Academy of the USA. †Number of tips in the phylogeny the OU model was fitted to. Where there were multiple analyses in a paper we use the median number of tips.  All packages mentioned are R packages apart from BayesTraits, COMPARE, MATLAB, Mesquite, PAM, PDAP and PDTREE.}\\
\bigskip        
\textbf{Paper}  &   \textbf{Year}    &   \textbf{Journal} &   \textbf{Taxon}   &   \textbf{Ntips}   &   \textbf{Use in paper}   &   \textbf{Stats/R package} \\

@@

Gvo\v{z}d\'{i}k and Damme 2006    &   2006    &   Evolution   &   amphibians  &   10  &   phylogenetic correction &   COMPARE \\

@@

Gomez and Th\'{e}ry 2007  &   2007    &   Am Nat  &   birds   &   40  &   model of evolution  &   OUCH    \\

@@

Arbour and L\'{o}pez-Fern\'{a}ndez 2013   &   2013    &   PRSB    &   fish    &   27  &   model of evolution  &   OUCH    \\

@@

Fr\'{e}d\'{e}rich et al. 2013 &   2013    &   Am Nat  &   fish    &   208 &   model of evolution  &   OUwie   \\

@@

L\'{o}pez-Fern\'{a}ndez et al. 2013   &   2013    &   Evolution   &   fish    &   135 &   model of evolution  &   GEIGER  \\

@@

P\'{e}rez i de Lanuza et al. 2013 &   2013    &   JEB &   reptiles    &   42  &   ancestral state reconstruction  &   GEIGER  \\

@@

Arbour, J. H., and H. L\'{o}pez-Fern\'{a}ndez. 2013. Ecological variation in South American geophagine cichlids arose during an early burst of adaptive morphological and functional evolution. Proceedings of the Royal Society B: Biological Sciences 280.\\

@@

Gomez, D., and M. Th\'{e}ry. 2007. Simultaneous crypsis and conspicuousness in color patterns: comparative analysis of a neotropical rainforest bird community. The American Naturalist 169:S42-S61.\\

@@

Gvo\v{z}d\'{i}k, L., and R. V. Damme. 2006. Triturus newts defy the running-swimming dilemma. Evolution 60:2110-2121.\\

@@

Huey, R. B., C. A. Deutsch, J. J. Tewksbury, L. J. Vitt, P. E. Hertz, H. J. \'{A}lvarez P\'{e}rez, and T. Garland. 2009. Why tropical forest lizards are vulnerable to climate warming. Proceedings of the Royal Society of London Series B-Biological Sciences 276:1939-1948.\\

@@

L\'{o}pez-Fern\'{a}ndez, H., J. H. Arbour, K. O. Winemiller, and R. L. Honeycutt. 2013. Testing for ancient adaptive radiations in Neotropical cichlid fishes. Evolution 67:1321-1337.\\

@@

Smith, S. D., C. An\'{e}, and D. A. Baum. 2008. The role of pollinator shifts in the floral diversification of Iochroma (Solanaceae). Evolution 62:793-806.\\

@@

Voje, K. L., A. B. Mazzarella, T. F. Hansen, K. \o{O}stbye, T. Klepaker, A. Bass, A. Herland, K. M. Bærum, F. Gregersen, and L. A. V\o{o}llestad. 2013. Adaptation and constraint in a stickleback radiation. Journal of evolutionary biology 26:2396-2414.\\

@@
